Bayes Computational Methods

### 2020-04-08 Paper.Rmd
paper edits

@@

>>>>

# Example Model

To illustrate how each bayesian computational method is used in practice, we will demonstrate how to construct a sampler for a simple Beta-Binomial model for each of the five methods we are studying.

## Data Preparation

First, we need to prepare the data that we will be using.

```{r}
p <- .25 # binomial p
n <-  10 # binomial n

set.seed(1)

(y <- rbinom(1, n, p))

model_data <- list(
  "y" = y,
  "N" = n
)
```

## Model Configuration

Next, we will pick some basic model parameters. Here, we will use four chains, a warmup of 1,000 iterations, and a total iteration count of 10,000.

```{r}
n_chains <- 4L
n_iter <- 1e4L
n_warmup <- 1e3L
```

## OpenBUGS / R2OpenBUGS

When using OpenBUGS through R, we will use the `R2OpenBUGS` library in order to pipe the data and model code into the program rather than using manual entry. This requires the creation of a temporary file that holds the model text that will be used.

```{r}
## load library
library(R2OpenBUGS)

## model script code
bugs_model <- function() {
  y ~ dbin(p,N)
  p ~ dbeta(1,1)
}

## parameters to monitor
bugs_monitor <- "p"

## temporary model script text file creation
bugs.file <- file.path(tempdir(), "model.txt")
write.model(bugs_model, bugs.file)

## run the model
bugs_fit <- bugs(
    "model.file" = bugs.file, "data" = model_data, "parameters.to.save" = bugs_monitor, 
    "inits" = NULL, "n.chains" = n_chains, "n.iter" = n_iter, "n.burnin" = n_warmup
)
```

In JAGS, the model script simply needs to be inputted as a string to the `run.jags()` function.
```{r}
## load libraries
library("rjags"); library("runjags")

## model script code
jags_model <- "
  model{
    y ~ dbin(p,n)
    p ~ dbeta(1,1)
  }
"

## parameters to monitor
jags_monitor <- c("p")

## run the model
jags_fit <- run.jags(
    "model" = jags_model, "data" = model_data, "monitor" = jags_monitor, 
    "n.chains" = n_chains, "sample" = n_iter, "burnin" = n_warmup
) 
```

An additional step needed when using Nimble, which is to specify initial values for all parameters of interest, since the sampler cannot generate initial values on its own.
```{r}
## load library
library(nimble)

## model script code
nimble_model <- nimbleCode({
  y ~ dbin(p,n)
  p ~ dbeta(1,1)
})

## parameters to monitor
nimble_monitor <- c("p")

## specify initial values of parameters
nimble_inits <- list(
  "p" = rbeta(1,1,1)
)

## run the model
nimble_fit <- nimbleMCMC(
    "code" = nimble_model, "data" = model_data, 
    "inits" = nimble_inits, "monitors" = nimble_monitor, "nchains" = n_chains, 
    "niter" = n_iter, "nburnin" = n_warmup, "summary" = TRUE
)
```

Running models in STAN requires a .stan file which contains the model script code, written in the unique STAN syntax. This can either be a seperate file, or can be inputted as a character string. It is generally recommended to use a seperate .stan file, espcially when using RStudio, which can do syntax highlighting and point out errors in STAN code.

Here is the .stan file for this example, which is saved as "stan_file.stan". Each STAN file contains a data block, parameters block, and model block at a minimum.
```{r}
data {
  int<lower=0> y;
  int<lower=0> n;
}

parameters {
  real p;                
}

model {
  y ~ binomial(n,p);
  p ~ beta(1,1);
}
```

The sampler can then be run simply using the `stan()` function call, with the .stan file location specified.

```{r}
## load library
library("rstan")

## run the model
stan_fit <- stan(
    "file" = "stan_file.stan", "data" = model_data, 
    "chains" = n_chains, "iter" = n_iter, "warmup" = n_warmup, 
    "control" = list("adapt_delta" = 0.99)
)
```

Each data variable needs to be converted to a greta array before it can be used in the model.
```{r}
## convert data into greta arrays
y <- greta::as_data(y)
n <- greta::as_data(n)
```

Next, the priors for each parameter of interested are defined using functions that describe the prior distribution paramter values. The distribution of the response variable given in the data is then described using the `distribution()` function. The model is compiled using the `model()` function, before being sampled by `mcmc()`.

```{r}
## load library
library(greta)

## specify prior for parameter(s) of interest
p <- beta(1,1)

## define distribution of data variable and relationship with parameter
distribution(y) <- binomial(n,p)

## compile model and run
greta_model <- model(p)
greta_fit <- mcmc(
    "model" = greta_model, "n_samples" = n_iter,
    "warmup" = n_warmup,"chains" = n_chains
)
```

@@

## STAN Probabilty and Sampling Statements

Inside the model block of a stan file, probability statements can be written in the same convention as in other bayesian sampling methods

```{r}
model{
  y ~ binomial(n,p);
  p ~ beta(1,1);
}
```

This kind of probability statement does not actually perform any sampling, but is instead tranformed into a statement about the incremental log probability density of a target distribution. The log probability density function can be specified directly.

```{r}
model {
  target += binomial_lpdf(y | n, p);
  target += beta_lpdf(p | 1, 1);
}
```

Either method yields the same result, but in some cases it is advantageous to use the incremental log probability density (or cumulative density) function to get exact log probability values for the model. In general, using the standard sampling statement syntax is preferred since it is easier to understand.
